---
title: "Assignment3-Airline Delay"
author: "Hui (Gracie) Han"
---

## Objective
(1) Create a .CSV file (or optionally, a MySQL database!) that includes all of the information above. You're encouraged to use a "wide" structure similar to how the information appears above, so that you can practice tidying and transformations as described below.
(2) Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy and transform your data.

# load needed package
```{r, warning=FALSE, message=FALSE}
library ('tidyr')
library ('dplyr')
library ('knitr')
```

###Get the data from the CSV File, the CSV file format is exactly the same as the source

```{r}
airlines <- read.csv ("E:/airlineData.csv")
kable(airlines)

kable(airlines)
```

## Data Cleaning to make it into a condition ready for analysis

```{r, comment=NA}
# rename cols
names(airlines)[c(1,2)] <- c("Airline", "Status")
airlines

# remove middle NA obs, in row3, and row6
airlines <- airlines [c(-3,-6),]
airlines

# add in missing airline values
airlines[2,1] <- "ALASKA"
airlines[4,1] <- "AMWEST"
kable (airlines)

# the original data is in wide format, make it a long format for cities
tidyairline1 <- airlines %>% 
   gather(cities, count, 3:7)
tidyairline1

#  then, for the status (on time vs delay, we transform it a wide format)
tidyairline2 <- tidyairline1 %>% 
     spread(Status, count)
 tidyairline2


#rename the 4th column name to get rid of the space in col name
names(tidyairline2)[4] <- "ontime"

kable(tidyairline2)

```
(3) Perform analysis to compare the arrival delays for the two airlines.
```{r}
#Delay ratios by airline and city, looked at each element seperately

tidyairline3 <- tidyairline2 %>%
   mutate(total=ontime + delayed, delayratio=round(delayed/total, 2)) 
tidyairline3


#Plot these delay ratios, based on City
dotchart(tidyairline3$delayratio, groups = tidyairline3$Airline, labels = tidyairline3$cities )
```
```{r}

```

## Observation from the analysis: 
It looks like ALASKA airlines had a lower individual delay ratio than AMWEST in every single city. 

